Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2021-12-14 src/ms.tex
addressing coauthor comments

In the CKS sample, there appears to be a clustering of stars above the ridge with \teff~$>6100$~K (seen most clearly in the top panels of Figure~\ref{fig:models}). This cluster of points has a similar slope in the \teff--\prot plane as the long-period pile-up, and does not reside on the harmonic of the ridge line as one might expect if the periods were erroneously determined. After inspecting the light curves of stars on the ridge and above the ridge in the temperature range 6100~K~$<$~ \teff~$<6300$~K we did not notice any obvious trends or differences between the two samples. The periods of the stars in the cluster above the ridge appear to be just as reliable as those on the ridge. Interestingly, a similar clustering of points is not observed in the LAMOST--Kepler sample and is less pronounced or absent when substituting the CKS temperatures with \teff from either \citet{Brewer2018} or \citet{Martinez2019}, two studies that independently derived spectroscopic parameters from the CKS spectra.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{./figures/asteroseismic.pdf}
    \caption{Comparison of CKS (left) and LAMOST--Kepler samples (right) with the \citet{Hall2021} main-sequence asteroseismic sample in the \teff--\prot plane. Note, the CKS, LAMOST, and \citet{Hall2021} samples derive \teff from distinct pipelines. Using LAMOST \teff, where it exists, for the \citet{Hall2021} sample brings that sample into closer agreement with the long-period pile-up in the LAMOST-McQuillan sample.}
    \label{fig:asteroseismic}
\end{figure}

@@

\citet{Curtis2019a, Curtis2020} proposed that such an effect could be produced by a temporary epoch of stalled spin-down, caused when the core transfers angular momentum to the envelope and offsets the effect of magnetic braking \citep[e.g.][]{MacGregor1991}. Angular momentum is exchanged between the envelope and the core on a characteristic timescale known as the core–envelope coupling timescale. Theoretical predictions for this timescale are mass-dependent, and range from 30--110 Myr for a solar-mass star \citep{Bouvier2008, IrwinBouvier2009, Denissenkov2010, GalletBouvier2015, Spada2020}. In the core–envelope coupling scenario, wind-driven spin-down stalls when the star's age becomes comparable to the core–envelope coupling timescale. Stars then resume spinning down via magnetic braking, and from open cluster data \citet{Curtis2020} estimated that solar-mass stars resume spin-down after $\approx$~230~Myr. This timescale is consistent with independent studies of open clusters, which show multiple cluster sequences of Sun-like stars overlapping at ages of $\sim$100--200~Myr, with spin-down resuming by $\sim$300~Myr \citep{Fritzewski2020,Fritzewski2021}.

@@

We emphasize that the temporary epoch of stalled magnetic braking proposed by \citet{Curtis2019a, Curtis2020} is not to be confused with the termination of magnetic braking that characterizes the WMB model of \citet{vanSaders2016, vanSaders2019}. The physical mechanisms thought to be responsible for each of these proposed stages of rotational evolution are distinct. As mentioned above, the leading theory for the earlier stage of stalled braking is core–envelope coupling. Crucially, in the stalled braking phase, wind-driven angular momentum losses are not ceased but rather offset by the spin-up torque from core–envelope coupling. In contrast, in the WMB scenario, wind-driven angular momentum losses are proposed to cease ($dJ/dt$=0), with subsequent rotational evolution dictated by the changes in the moment of inertia. The physical mechanism responsible for WMB is not known, but one hypothesis is that the efficiency of magnetic braking declines with magnetic field complexity \citep[e.g.][]{Reville2015, vanSaders2016, Garraffo2016, Metcalfe2016, Metcalfe2019}.

@@

An unexplained feature of the Kepler rotation period distribution is the existence of a bimodal period distribution for stars of similar \teff. The effect was first noticed for M-dwarfs \citep{McQuillan2013b}, but was later shown to extend to $\sim$5000~K \citep{Reinhold2013, McQuillan2014, ReinholdHekker2020}, and even to $\sim$6500~K \citep{Davenport2017}. \citet{McQuillan2013b, McQuillan2014} speculated that this period bimodality could originate from stellar populations of different ages, an explanation seemingly supported by a correlation between the strength of the bimodality and height above the galactic disk \citep{DavenportCovey2018}. However, \citet{Reinhold2019} found that the dearth of stars with intermediate rotation periods is associated with a decrease in photometric variability. Consequently, those authors proposed that the period bimodality may be the result of a transition between spot- and faculae-dominated photospheres. In this scenario, the period gap is due to bright faculae canceling out the effects of dark star spots.

The short- and long-period pile-ups we examine here naturally produce a dearth of rotators at intermediate periods. This gap is the same period gap noticed by the authors mentioned above, as made apparent when comparing the \lamostmcq sample to the original \citet{McQuillan2014} sample. Moreover, as seen in Figure~\ref{fig:gap}, we recover the gradient in photometric variability across the gap pointed out by \citet{Reinhold2019}. The photometric variability, as measured through the \rper metric published by \citet{McQuillan2014}, reaches a local minimum at the location of the gap. This supports the notion that the cause of the gap is due to changes in the stars themselves, rather than being the result of mixed stellar populations. However, the variability levels on both sides of the gap are not close to the detection limit, as evidenced by the fact that periods are securely detected for stars with similar properties at much lower \rper values. This would suggest that the period gap is not due solely to a detection issue, unless variability levels were to drop precipitously as stars approached the gap.

>>>>

In the CKS sample, there appears to be a clustering of stars above the ridge with \teff~$>6100$~K (seen most clearly in the top panels of Figure~\ref{fig:models}). This cluster of points has a similar slope in the \teff--\prot plane as the long-period pile-up, and does not reside on the harmonic of the ridge line as one might expect if the periods were erroneously determined. After inspecting the light curves of stars on the ridge and above the ridge in the temperature range 6100~K~$<$~ \teff~$<6300$~K we did not notice any obvious trends or differences between the two samples. The periods of the stars in the cluster above the ridge appear to be just as reliable as those on the ridge. Interestingly, a similar clustering of points is not observed in the LAMOST--Kepler sample and is less pronounced or absent when substituting the CKS temperatures with \teff from either \citet{Brewer2018} or \citet{Martinez2019}, two studies that independently derived spectroscopic parameters from the CKS spectra. Further inspection of the stars in this cluster revealed that they have anomalously large \teff discrepancies between the CKS and SPOCS catalogs, such that the SPOCS temperatures shift the stars onto the long-period pile-up. We conclude that the temperatures for these stars in the CKS catalog are too high by $\gtrsim$~100~K.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{./figures/asteroseismic.pdf}
    \caption{Comparison of CKS (left) and LAMOST--Kepler samples (right) with the \citet{Hall2021} main-sequence asteroseismic sample in the \teff--\prot plane. Note, the CKS, LAMOST, and \citet{Hall2021} samples derive \teff from distinct pipelines. A constant offset of +41~K was applied to the LAMOST \teff to bring that sample onto the same scale as the asteroseismic sample (see Appendix~\ref{app:teff}). Using LAMOST \teff, where it exists, for the \citet{Hall2021} sample brings that sample into even closer agreement with the long-period pile-up in the LAMOST-McQuillan sample.}
    \label{fig:asteroseismic}
\end{figure}

@@

\citet{Curtis2019a, Curtis2020} proposed that such an effect could be produced by a temporary epoch of stalled spin-down, caused either by (a) core–envelope momentum transfer which offsets the effect of magnetic braking \citep[e.g.][]{MacGregor1991}, or (b) a reduction in the magnetic braking torque. Angular momentum is exchanged between the envelope and the core on a characteristic timescale known as the core–envelope coupling timescale. Theoretical predictions for this timescale are mass-dependent, and range from 30--110 Myr for a solar-mass star \citep{Bouvier2008, IrwinBouvier2009, Denissenkov2010, GalletBouvier2015, Spada2020}. In the core–envelope coupling scenario, wind-driven spin-down stalls when the star's age becomes comparable to the core–envelope coupling timescale. Stars then resume spinning down via magnetic braking, and from open cluster data \citet{Curtis2020} estimated that solar-mass stars resume spin-down after $\approx$~230~Myr. This timescale is consistent with independent studies of open clusters, which show multiple cluster sequences of Sun-like stars overlapping at ages of $\sim$100--200~Myr, with spin-down resuming by $\sim$300~Myr \citep{Fritzewski2020,Fritzewski2021}.

@@

We emphasize that the temporary epoch of stalled magnetic braking proposed by \citet{Curtis2019a, Curtis2020} is not to be confused with the termination of magnetic braking that characterizes the WMB model of \citet{vanSaders2016, vanSaders2019}. The physical mechanisms thought to be responsible for each of these proposed stages of rotational evolution are distinct. As mentioned above, one theory for the earlier stage of stalled braking is core–envelope coupling. Crucially, in the stalled braking phase, wind-driven angular momentum losses are not ceased but rather offset by the spin-up torque from core–envelope coupling. In contrast, in the WMB scenario, wind-driven angular momentum losses are proposed to cease ($dJ/dt$=0), with subsequent rotational evolution dictated by the changes in the moment of inertia. The physical mechanism responsible for WMB is not known, but one hypothesis is that the efficiency of magnetic braking declines with magnetic field complexity \citep[e.g.][]{Reville2015, vanSaders2016, Garraffo2016, Metcalfe2016, Metcalfe2019}.

@@

An unexplained feature of the Kepler rotation period distribution is the existence of a bimodal period distribution for stars of similar \teff. The effect was first noticed for M-dwarfs \citep{McQuillan2013b}, but was later shown to extend to $\sim$5000~K \citep{Reinhold2013, McQuillan2014, ReinholdHekker2020}, and even to $\sim$6500~K \citep{Davenport2017}. \citet{McQuillan2013b, McQuillan2014} speculated that this period bimodality could originate from stellar populations of different ages, an explanation seemingly supported by a correlation between the strength of the bimodality and height above the galactic disk \citep{DavenportCovey2018}. However, \citet{Curtis2020} demonstrated that cluster sequences cross the gap, invalidating the claim that the feature is caused at a specific age, as one might expect from a period of decreased star formation. Additionally, \citet{Gordon2021} found that the gap is observed across the many fields observed by the K2 mission, further invalidating the star formation history hypothesis as different Galactic sight lines are expected to have different different star formation histories. An alternative explanation for the gap was proposed by \citet{Reinhold2019}, who found that the dearth of stars with intermediate rotation periods is associated with a decrease in photometric variability. Consequently, those authors proposed that the period bimodality may be the result of a transition between spot- and faculae-dominated photospheres. In this scenario, the period gap is due to bright faculae canceling out the effects of dark star spots.

The short- and long-period pile-ups we examine here naturally produce a dearth of rotators at intermediate periods. This gap is the same period gap noticed by the authors mentioned above, as made apparent when comparing the \lamostmcq sample to the original \citet{McQuillan2014} sample. Moreover, as seen in Figure~\ref{fig:gap}, we recover the gradient in photometric variability across the gap pointed out by \citet{Reinhold2019}. The photometric variability, as measured through the \rper metric published by \citet{McQuillan2014}, reaches a local minimum near the location of the gap. This supports the notion that the cause of the gap is due to changes in the stars themselves, rather than being the result of mixed stellar populations. However, the variability levels on both sides of the gap are not close to the detection limit, as evidenced by the fact that periods are securely detected for stars with similar properties at much lower \rper values. This would suggest that the period gap is not due solely to a detection issue, unless variability levels were to drop precipitously as stars approached the gap.
